---
layout: lesson
title: Lesson 21
subtitle: Metaphors for Bitcoin's future
quote: "I know something interesting is sure to happen..."
categories: [bitcoin, lesson]
audio: /assets/audio/21lessons/3-21.m4a
---

In the last couple of decades, it became apparent that technological
innovation does not follow a linear trend. Whether you believe in the
technological singularity or not, it is undeniable that progress is
exponential in many fields. Not only that, but the rate at which
technologies are being adopted is accelerating, and before you know it
the bush in the local schoolyard is gone and your kids are using
Snapchat instead. Exponential curves have the tendency to slap you in
the face way before you see them coming.

Bitcoin is an exponential technology built upon exponential
technologies. [Our World in Data] beautifully shows [the rising speed of
technological adoption], starting in 1903 with the introduction of
landlines. Landlines, electricity, computers, the internet, smartphones;
all follow exponential trends in price-performance and adoption. Bitcoin
does too.

{% include image.html otherPost="2019-04-02-technological-teachings-of-bitcoin" name="tech-adoption.png" caption="Bitcoin is literally off the charts." %}

Bitcoin has not one but [multiple network effects], all of which
resulting in exponential growth patterns in their respective area:
price, users, security, developers, market share, and adoption as global
money.

Having survived its infancy, Bitcoin is continuing to grow every day in
more aspects than one. Granted, the technology has not reached maturity
yet. It might be in its adolescence. But if the technology is
exponential, the path from obscurity to ubiquity is short.

{% include image.html otherPost="2019-04-02-technological-teachings-of-bitcoin" name="mobile-phone.png" caption="Mobile phone, ca 1965 vs 2019." %}

In his 2003 [TED talk], Jeff Bezos chose to use electricity as a
metaphor for the web's future. All three phenomena --- electricity, the
internet, Bitcoin --- are *enabling* technologies, networks which enable
other things. They are infrastructure to be built upon, foundational in
nature.

Electricity has been around for a while now. We take it for granted. The
internet is quite a bit younger, but most people already take it for
granted as well. Bitcoin is ten years old and has entered public
consciousness during the last hype cycle. Only the earliest of adopters
take it for granted. As [more time] passes, more and more people will
recognize Bitcoin as something which simply is.

In 1994, the internet was still confusing and unintuitive. Watching this
old [recording of the Today Show] makes it obvious that what feels
natural and intuitive now actually wasn't back then. Bitcoin is still
confusing and alien to most, but just like the internet is second nature
for digital natives, spending and [stacking] sats will be second nature
to the bitcoin natives of the future.

> "The future is already here --- it's just not very evenly
> distributed."
> <cite>[William Gibson]</cite>

In 1995, about 15% of American adults used the internet. Historical
[data from the Pew Research Center] shows how the internet has woven
itself into all our lives. According to a [consumer survey] by Kaspersky
Lab, 13% of respondents have used Bitcoin and its clones to pay for
goods in 2018. While payments aren't the only use-case of bitcoin, it is
some indication of where we are in Internet time: in the early- to
mid-90s.

In 1997, Jeff Bezos stated in a [letter to shareholders] that "this is
Day 1 for the Internet," recognizing the great untapped potential for
the internet and, by extension, his company. Whatever day this is for
Bitcoin, the vast amounts of untapped potential are clear to all but the
most casual observer.

{% include image.html otherPost="2019-04-02-technological-teachings-of-bitcoin" name="internet-evolution-black-dates.png" caption="The internet, 1982 vs 2005. Source: cc-by Merit Network, Inc. and Barrett Lyon, Opte Project" %}


Bitcoin's first node went online in 2009 after Satoshi mined the
[genesis block] and released the software into the wild. His node wasn't
alone for long. Hal Finney was one of the first people to pick up on the
idea and join the network. Ten years later, as of this writing, more
than
[75.000](https://luke.dashjr.org/programs/bitcoin/files/charts/software.html)
nodes are [running bitcoin].

The protocol's base layer isn't the only thing growing exponentially.
The lightning network, a second layer technology, is growing at an even
faster rate.

In January 2018, the lightning network had [40 nodes] and 60 channels.
In April 2019, the network grew to more than 4000 nodes and around
40.000 channels. Keep in mind that this is still experimental technology
where loss of funds can and does occur. Yet the [trend is clear][Jameson Lopp]:
thousands of people are [reckless] and eager to use it.

{% include image.html otherPost="2019-04-02-technological-teachings-of-bitcoin" name="lnd-growth-lopp-black.png" caption="Lightning Network, January 2018 vs December 2018. Source: Jameson Lopp" %}

To me, having lived through the meteoric rise of the web, the parallels
between the internet and Bitcoin are obvious. Both are networks, both
are exponential technologies, and both enable new possibilities, new
industries, new ways of life. Just like electricity was the best
metaphor to understand where the internet is heading, the internet might
be the best metaphor to understand where bitcoin is heading. Or, in the
words of Andreas Antonopoulos, Bitcoin is [*The Internet of Money*].
These metaphors are a great reminder that while history doesn't repeat
itself, it often rhymes.

Exponential technologies are hard to grasp and often underestimated.
Even though I have a great interest in such technologies, I am
constantly surprised by the pace of progress and innovation. Watching
the Bitcoin ecosystem grow is like watching the rise of the internet in
fast-forward. It is exhilarating.

My quest of trying to make sense of Bitcoin has led me down the pathways
of history in more ways than one. Understanding ancient societal
structures, past monies, and how communication networks evolved were all
part of the journey. From the handaxe to the smartphone, technology has
undoubtedly changed our world many times over. Networked technologies
are especially transformational: writing, roads, electricity, the
internet. All of them changed the world. Bitcoin has changed mine and
will continue to change the minds and hearts of those who dare to use
it.

Bitcoin taught me that understanding the past is essential to
understanding its future. A future which is just beginning.

---

#### Down the Rabbit Hole

- [The Rising Speed of Technological Adoption][the rising speed of technological adoption] by Jeff Desjardins
- [The 7 Network Effects of Bitcoin][multiple network effects] by Trace Mayer
- [The Electricity Metaphor for the Web's Future][TED talk] by Jeff Bezos
- [How the internet has woven itself into American life][data from the Pew Research Center] by Susannah Fox and Lee Rainie
- [Genesis Block][genesis block] on the Bitcoin Wiki
- [Lindy Effect][more time] on Wikipedia

[Our World in Data]: https://ourworldindata.org/
[the rising speed of technological adoption]: https://www.visualcapitalist.com/rising-speed-technological-adoption/
[multiple network effects]: https://www.thrivenotes.com/the-7-network-effects-of-bitcoin/
[TED talk]: https://www.ted.com/talks/jeff_bezos_on_the_next_web_innovation
[recording of the Today Show]: https://www.youtube.com/watch?v=UlJku_CSyNg
[William Gibson]: https://www.npr.org/2018/10/22/1067220/the-science-in-science-fiction
[data from the Pew Research Center]: https://www.pewinternet.org/2014/02/27/part-1-how-the-internet-has-woven-itself-into-american-life/
[consumer survey]: https://www.kaspersky.com/blog/money-report-2018/
[letter to shareholders]: http://media.corporate-ir.net/media_files/irol/97/97664/reports/Shareholderletter97.pdf
[running bitcoin]: https://twitter.com/halfin/status/1110302988?lang=en
[40 nodes]: https://bitcoinist.com/bitcoin-lightning-network-mainnet-nodes/
[reckless]: https://twitter.com/hashtag/reckless
[Jameson Lopp]: https://twitter.com/lopp/status/1077200836072296449
[*The Internet of Money*]: https://theinternetofmoney.info/
[stacking]: https://twitter.com/hashtag/stackingsats

<!-- Bitcoin Wiki -->
[genesis block]: https://en.bitcoin.it/wiki/Genesis_block

<!-- Wikipedia -->
[more time]: https://en.wikipedia.org/wiki/Lindy_effect
[alice]: https://en.wikipedia.org/wiki/Alice%27s_Adventures_in_Wonderland
[carroll]: https://en.wikipedia.org/wiki/Lewis_Carroll
