Falling down the Bitcoin rabbit hole is a strange experience. Like many others,
I feel like I have learned more in the last couple of years studying Bitcoin
than I have during two decades of formal education.

The following lessons are a distillation of what I’ve learned. First published
as an article series titled [“What I’ve Learned From Bitcoin,”][I] what follows
can be seen as a second edition of the original series.

Like Bitcoin, these lessons aren't a static thing. I plan to work on them
periodically, releasing updated versions and additional material in the future.

Unlike Bitcoin, future versions of this project do not have to be backward
compatible. Some lessons might be extended, others might be reworked or
replaced. I hope that a future version will be something you can hold in your
hands, but I don’t want to promise anything just yet.

Bitcoin is an inexhaustible teacher, which is why I do not claim that these
lessons are all-encompassing or complete. They are a reflection of my personal
journey down the rabbit hole. There are many more lessons to be learned, and
every person will learn something different from entering the world of Bitcoin.

I hope that you will find these lessons useful and that the process of learning
them by reading won’t be as arduous and painful as learning them firsthand.

[Gigi][dergigi]

<!-- Internal -->
[I]: {% post_url /bitcoin/2018-12-21-philosophical-teachings-of-bitcoin %}

<!-- Twitter -->
[dergigi]: https://twitter.com/dergigi

<!-- Wikipedia -->
[alice]: https://en.wikipedia.org/wiki/Alice%27s_Adventures_in_Wonderland
[carroll]: https://en.wikipedia.org/wiki/Lewis_Carroll
