\chapter*{Foreword to the Swan Edition}
\pdfbookmark{Swan Edition}{swan-edition}

Arguably, Bitcoin is still one of the most widely misunderstood phenomena of our
time. It is hard to blame anyone for this. The world is a complicated place,
Bitcoin is a mysterious beast, and people, in general, are busy and don't have
the time to worry about Magic Internet Money.

The majority of the population still believes that Bitcoin can't work, or if it
works, governments will shut it down. Or, if it works and isn't shut down, they
fear that they have missed the boat. Bitcoin, however, isn't a boat. It is the
water that all other boats rest upon. Understanding and appreciating this fact
will allow you to capitalize on the global monetary revolution currently
underway.

But there is more to be had than monetary gains. Bitcoin is a liberating force,
and this liberation is contagious. With Bitcoin, everyone has the freedom to
save, the freedom to transact, and the freedom to spend their financial energy
as they see fit. Due to the economic and political instability of current times,
these freedoms will only grow in importance. In a world where tech oligarchies
dominate, and monetary institutions are supposedly too big to fail, every step
towards more independence and self-sovereignty is a step in the right direction.

It is undeniable that the fabric of society is shifting before our eyes. While
the reasons for these shifts are manifold, the fact that our money - the glue
that holds a complex society together - is being watered down doesn't help. The
certainty of Bitcoin's rule set brings a fresh wind of truthfulness to a world
burdened by deceit and broken promises. And with this wind comes
\href{https://hope.com}{hope}.

Because of these profound societal shifts, I am convinced that investing time
into gaining a deeper understanding of Bitcoin is the most important investment
that you can make for your future self. The second most important investment
will follow naturally after that: getting your hands on some bitcoin.

Despite crazy price rallies and steep drops that go along with them, the
worldwide adoption of a Bitcoin Standard is a marathon, not a sprint. However,
thanks to services like \href{https://swanbitcoin.com}{Swan}, you have the
option to front-run the rest of the world and adopt the Bitcoin Standard for
yourself, on your own terms. I believe that committing to an auto-DCA plan and
starting to accumulate the best money the world has ever seen is the best
strategy to get acquainted with these strange waters. Arguably, it is also the
best way to smoothen the waves of a choppy sea.

This book only scratches the surface of the orange juggernaut. However, I hope
that it will serve as a guide, pointing towards the many entry points of the
Bitcoin rabbit hole. Like everyone else, I am
\href{https://twitter.com/dergigi/status/1285485658656649216?s=20}{still
exploring} this strange beast, trying to make sense of it all. And I hope that,
after reading \textit{21 Lessons}, some parts of it will make a little bit more
sense to you as well.

Gigi

\begin{flushright}
  Gigi \\
  January 15, 2021
\end{flushright}
