\chapter*{Acknowledgments}
\pdfbookmark{Acknowledgments}{acknowledgments}

Thanks to the countless authors and content producers who influenced my thinking
on Bitcoin and the topics it touches. There are too many to list them all, but
I’ll do my best to name a few.

\begin{itemize}
  \item Thanks to Arjun Balaji for the tweet which motivated me to write this.
  \item Thanks to Marty Bent for providing endless food for thought and entertainment. If you are not subscribed to Marty’s Bent and Tales From The Crypt, you are missing out. Cheers Matt and Marty for guiding us through the rabbit hole.
  \item Thanks to Michael Goldstein and Pierre Rochard for curating and providing the greatest Bitcoin literature via the Nakamoto Institute. And thank you for creating the Noded Podcast which influenced my philosophical views on Bitcoin substantially.
  \item Thanks to Peter McCormack for his honest tweets and the What Bitcoin Did podcast, which keeps providing great insights from many areas of the space.
  \item Thanks to Andreas M. Antonopoulos for all the educational material he has put out over the years.
  \item Thanks to Saifedean Ammous for his convictions, savage tweets, and writing The Bitcoin Standard
  \item Thanks to Francis Pouliot for sharing his excitement about finding out about the timechain.
  \item Thanks to Jannik, Brandon, Matt, Camilo, Daniel, Michael, and Raphael for providing feedback to early drafts of some lessons. Special thanks to Jannik who proofread multiple drafts multiple times.
  \item Thanks to Dhruv Bansal and Matt Odell for taking the time to discuss some of these ideas with me.
  \item Thanks to Guy Swann for producing an audio version of 21lessons.com.
  \item Thanks to my wife for putting up with me and my obsessive nature.
  \item Thanks to my family for supporting me during both the good times and the bad.
  \item Last but not least, thanks to all the bitcoin maximalists, shitcoin minimalists, shills, bots, and shitposters which reside in the beautiful garden that is Bitcoin twitter.
\end{itemize}

And finally, thank you for reading this. I hope you enjoyed it as much as I did enjoy writing it.
