\chapter*{What Now?}
\pdfbookmark{What Now?}{what-now}

Getting into Bitcoin can be overwhelming in many ways. Over the years, all kinds
of people reached out to me to ask the ever-so-tricky question: What now?

The question is impossible to answer, of course. Situations are different, risk
appetites are varied, and what might be sensible to one person seems like
insanity to the next. However, as far as
\href{https://dergigi.com/2020/04/27/dear-family-dear-friends/}{my friends and
family} are concerned, I tell them one thing above all else: "start stackings
sats, today."

Dipping your toes in the water - getting some skin in the game - is essential to
further your understanding. I believe that starting small and learning over time
is the most sensible approach. It is easy to shoot yourself in the foot. Better
to learn while what you're handling is still a small squirt gun and not a
full-blown bazooka.

Additionally, accumulating small bits of bitcoin is also an ingenious method to
deal with its inherent volatility. Dollar-cost-averaging allows you to stay sane
and get a small (but growing!) exposure to this alien bearer asset. And the
bearer asset part is important: if you don't hold the keys to your bitcoin, what
you are calling "your bitcoin" aren't actually your bitcoin.

Swan allows you to do both: (1) \href{https://www.swanbitcoin.com/}{set up an
automatic DCA plan}, and (2) automatically withdraw to your own wallet. As long
as there is fiat money to be spent, I believe this to be the best and easiest
strategy for most people. Start small, and go from there. And always keep
learning.
